---
title: 'Statistical Rethinking Problems: Chapter 2'
author: "Simon Thornewill von Essen"
date: "2024-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

## 2E1

Which of the expressions below correspond to the statement: *the probability of
rain on Monday*

$$P(\textrm{rain}) \\
P(\textrm{rain}|\textrm{Monday}) \\
P(\textrm{Monday}|\textrm{rain}) \\
\frac{P(\textrm{rain, Monday})}{P(\textrm{Monday})}$$

My answer is $P(\textrm{rain}|\textrm{Monday})$ because if something happen on
Monday then we are conditioning on it. The first one is the probability that
it rains without conditioning on anything, while the next is the probability 
it is Monday given that it rains.

The last one looks like a posterior distribution of some kind, but I don't think
it's what we're looking for.

## 2E2 

Which of the following statements corresponds to the expression 
$P(\textrm{Monday}|\textrm{rain})$

1. The probability of rain on Monday
2. The probability of rain, given that it is Monday
3. The probability that it is Monday, given that it is raining
4. The probability that it is Monday and that it is raining

The answer is 3, because it is the probability that it is Monday given that it
is raining. It cannot be 1 or 2 because it's not the probability that it is
raining to start with. It's also not a joint probability as in 4, which would
be $P(\textrm{rain}, \textrm{Monday})$.

## 2E3 

Skip...

## 2E4

The Bayesian statistician Bruno de Finetti began his 1973 book on probability
theory with the declaration "PROBABILITY DOES NOT EXIST." The capitals appeared
in the original, so I imagine de Finetti wanted us to shout this statement. 
What he meant is that probability is a device for describing uncertainty from
the perspective of an observer with limited knowledge; it has no objective 
reality. Discuss the globe tossing example from the chapter, in light of this
statement. What does it mean to say "the probability of water is 0.7"?

In order for something to be uncertain we simply don't have access to what the
true value for something is so by saying that the probability of water takes a
certain value we quantify that uncertainty using a parameter value. However,
we don't know if this parameter is true either and so we can try to encode that
uncertainty further in more complex ways.

## 2M1

Recall the globe tossing model from the chapter. Compute and plot the grid 
approximate posterior dist for each of the following sets of observations. In
each case, assume a uniform prior for p.

1. W, W, W
2. W, W, W, L
3. L, W, W, L, W, W, W

```{r}
data = c(1, 1, 1)

grid_prior = seq(0, 1-0.001, 0.001)
prior = rep(1, 1000)

grid.df = data.frame(params=grid_prior,
                     prior=prior)

grid.df["posterior_WWW"] = dbinom(sum(data), length(data), grid_prior)
grid.df["posterior_WWW"] = grid.df["posterior_WWW"]/sum(grid.df["posterior_WWW"])

head(grid.df)

```
We can now plot the posterior.

```{r}
ggplot(grid.df, aes(x = params, y = posterior_WWW)) +
  geom_line() +
  labs(x = "Params", y = "Posterior") +
  ggtitle("Posterior Plot")
```

Now we can do the same for if we were to draw an L, we can use the same 
posterior before as our prior but I calculated it from scratch because it was
easy enough.

```{r}
grid.df["posterior_WWWL"] = dbinom(3, 4, grid_prior)
grid.df["posterior_WWWL"] = grid.df["posterior_WWWL"] /
  sum(grid.df["posterior_WWWL"])

ggplot(grid.df, aes(x = params, y = posterior_WWWL)) +
  geom_line() +
  labs(x = "Params", y = "Posterior") +
  ggtitle("Posterior Plot")
```

So now for the final one we have 5 positives out of 7 trials...

```{r}
grid.df["posterior_WWWLWWL"] = dbinom(5, 7, grid_prior)
grid.df["posterior_WWWLWWL"] = grid.df["posterior_WWWLWWL"] /
  sum(grid.df["posterior_WWWL"])

ggplot(grid.df, aes(x = params, y = posterior_WWWLWWL)) +
  geom_line() +
  labs(x = "Params", y = "Posterior") +
  ggtitle("Posterior Plot")
```

```{r}
head(grid.df)
```

## 2M2

Now assume a prior for p that is equal to zero when p < 0.5 and is a positive 
constant when  p ≥ 0.5. Again compute and plot the grid approximate posterior
distribution for each of the sets of observations in the problem just above

```{r}
grid.df.new = grid.df
grid.df.new["prior"] = append(rep(0, 500), rep(1, 500))

grid.df.new["posterior_WWW"] = grid.df.new["prior"]*grid.df["posterior_WWW"]
grid.df.new["posterior_WWW"] = grid.df.new["posterior_WWW"]/
  sum(grid.df.new["posterior_WWW"])

grid.df.new["posterior_WWWL"] = grid.df.new["prior"]*grid.df["posterior_WWWL"]
grid.df.new["posterior_WWWL"] = grid.df.new["posterior_WWWL"]/
  sum(grid.df.new["posterior_WWWL"])

grid.df.new["posterior_WWWLWWL"] = grid.df.new["prior"]*grid.df["posterior_WWWLWWL"]
grid.df.new["posterior_WWWLWWL"] = grid.df.new["posterior_WWWLWWL"]/
  sum(grid.df.new["posterior_WWWLWWL"])

head(grid.df.new)
```

Now we can comvine the columns and plot

```{r}
grid.df.new.pivot = pivot_longer(grid.df.new, 
                            cols = starts_with("posterior"), 
                            names_to = "data", 
                            values_to = "posterior") %>%
  mutate(data = sub("^posterior_", "", data))

ggplot(grid.df.new.pivot, aes(x = params, y = posterior, color=data)) +
  geom_line() +
  labs(x = "Params", y = "Posterior") +
  ggtitle("Posterior Plot")

```

## 2M3

Suppose there are two globes, one for Earth and one for Mars. The Earth globe is 
70% covered  in water. The Mars globe is 100% land. Further suppose that one of 
these globes—you don’t know which— was tossed in the air and produced a “land” 
observation. Assume that each globe was equally  likely to be tossed. Show that 
the posterior probability that the globe was the Earth, conditional on seeing
"land" ($P(\textrm{Earth}|\textrm{land})$), is 0.23.

```{r}

df.planets = data.frame(
  planet = c("Earth", "Mars"),
  prior = rep(0.5, 2),
  p_land = c(0.3, 1)
) %>%
  mutate(likelihood=dbinom(1, 1, p_land),
         posterior=likelihood*prior/sum(likelihood*prior))

df.planets

```

We can see from the table that the posterior distribution that we have earth 
given that we saw land once (and the probabilities given) is 23%.

## 2M4

Suppose you have a deck with only three cards. Each card has two sides, and each 
side is either black or white. The cards are as follows:

1. One card has two black sides. 
2. The second card has one black and one white side. 
3. The  third card has two white sides. 

Now suppose all three cards are placed in a bag and shuffled. Someone  reaches 
into the bag and pulls out a card and places it flat on a table. A black side is 
shown facing up, but you don’t know the color of the side facing down. 

Show that the probability that the other side is  also black is 2/3. 

Hint: Use the counting method (Section 2 of the chapter) to approach this 
problem. This  means counting up the ways that each card could produce the 
observed data (a black side facing up on the table.

```{r}
df.cards = data.frame(
  card_side_1_is_black=c(1, 1, 0),
  card_side_2_is_black=c(1, 0, 0),
  prior=rep(1/3, 3)
) %>%
  mutate(ways=card_side_1_is_black + card_side_2_is_black,
         posterior=ways*prior/sum(ways*prior))

df.cards
```

We can see that the probability that we chose the double sided black card is 2/3
or or ~66%.

## 2M5 

Now suppose there are four cards: B/B, B/W, W/W, and another B/B. Again suppose 
a card is  drawn from the bag and a black side appears face up. Again calculate 
the probability that the other  side is black.

```{r}
df.cards = data.frame(
  card_side_1_is_black=c(1, 1, 0),
  card_side_2_is_black=c(1, 0, 0),
  prior=c(2/4, 1/4, 1/4)
) %>%
  mutate(ways=card_side_1_is_black + card_side_2_is_black,
         posterior=ways*prior/sum(ways*prior))

df.cards
```

this should be the same as below...

```{r}
df.cards = data.frame(
  card_side_1_is_black=c(1, 1, 0, 1),
  card_side_2_is_black=c(1, 0, 0, 1),
  prior=c(1/4, 1/4, 1/4, 1/4)
) %>%
  mutate(ways=card_side_1_is_black + card_side_2_is_black,
         posterior=ways*prior/sum(ways*prior))

df.cards %>%
  filter((card_side_1_is_black==1) & (card_side_2_is_black==1)) %>%
  summarise(posterior_prob=sum(posterior))
```

Indeed, we can see that the two methods are equivalent. 

## 2M6

Imagine that black ink is heavy, and so cards with black sides are heavier than 
cards with white  sides. As a result, it’s less likely that a card with black 
sides is pulled from the bag. 

So again assume there are three cards: B/B, B/W, and W/W. 

After experimenting a number of times, you conclude that for every way to pull 
the B/B card from the bag, there are 2 ways to pull the B/W card and 3 ways to  
pull the W/W card. Again suppose that a card is pulled and a black side appears 
face up. Show that  the probability the other side is black is now 0.5. Use the 
counting method, as before. 

```{r}
df.cards = data.frame(
  card_side_1_is_black=c(1, 1, 0),
  card_side_2_is_black=c(1, 0, 0),
  prior=c(1/6, 2/6, 3/6)
) %>%
  mutate(ways=card_side_1_is_black + card_side_2_is_black,
         posterior=ways*prior/sum(ways*prior))

df.cards
```

It's really interesting how the number of ways to pull the results actually gets
encoded into the prior, the odds we pull the double sided black card is 2:1.

## 2m7

Assume again the original card problem, with a single card showing a black side 
face up. 

Before looking at the other side, we draw another card from the bag and lay it 
face up on the table. The face that is shown on the new card is white. Show that 
the probability that the first card, the one showing a black side, has black on 
its other side is now 0.75. Use the counting method, if you can. 

Hint: Treat this like the sequence of globe tosses, counting all the ways to see 
each observation, for each possible first card.

It was difficult to represent this in code so the best way will be as follows

* black-black: 2 ways
  * white-black: 1 way
  * white-white: 2 ways
* black-white: 1 way
  * black-black: Not possible, 0
  * white-white: 2 way
* white-white: Not possible, 0

You then find the total number of ways we could have seen the data

```{r}
bb = 2
wb_bb = 1
ww_bb = 2
wb = 1
bb_wb = 0
ww_wb = 2

p_bb = bb * (wb_bb + ww_bb)

p_wb = wb * (ww_wb)

p_bb / (p_bb + p_wb)

```

So, I think my problem was that I was trying to add column together...


```{r}
cards <- c("Black-Black", "Black-White", "White-White")
combinations <- expand.grid(Card_Type1 = cards, 
                            Card_Type2 = cards) %>%
  filter(Card_Type1 != Card_Type2) %>%
  arrange(Card_Type1, Card_Type2) %>%
  mutate(Ways_Card1=c(2, 2, 1, 1, 0, 0),
         Ways_Card2=c(1, 2, 0, 2, 0, 1),
         prod=Ways_Card1*Ways_Card2,
         post=prod/sum(prod))

combinations %>%
  filter(Card_Type1=="Black-Black") %>%
  summarise(post_agg=sum(post))
```
## 2h1

Suppose there are two species of panda bear. Both are equally common in the wild 
and live  in the same places. They look exactly alike and eat the same food, and 
there is yet no genetic assay capable of telling them apart. They differ 
however in their family sizes. 

* Species A gives birth to twins 10% of the time, otherwise birthing a single 
infant. 
* Species B births twins 20% of the time, otherwise birthing singleton infants. 

Assume these numbers are known with certainty, from many years of field
research.

Now suppose you are managing a captive panda breeding program. You have a new 
female panda  of unknown species, and she has just given birth to twins. What is
the probability that her next birth will be twins.

I suppose that what we'd want to do is calculate the probability of the species

```{r}
df.pandas = data.frame(
  panda_type=c("A", "B"),
  prior=c(0.5, 0.5),
  likelihood=c(0.1, 0.2)
) %>%
  mutate(post=prior*likelihood/sum(prior*likelihood),
         next_post=c(0.1, 0.2),
         p_next_post=post*next_post)

df.pandas
```

We know that we have one set of twins, so it's just a matter of finding the 
likelihood of the next set of twins depending on the type

```{r}
sum(df.pandas["p_next_post"])
```

## 2h2 

Recall all the facts from the problem above. Now compute the probability that 
the panda we  have is from species A, assuming we have observed only the first 
birth and that it was twins.

We already know from the table above that it's 1/3.

## 2H3 

Continuing on from the previous problem, suppose the same panda mother has a 
second birth  and that it is not twins, but a singleton infant. Compute the 
posterior probability that this panda is  species A. 

```{r}
df.pandas = data.frame(
  panda_type=c("A", "B"),
  prior=c(1/3, 2/3),
  likelihood=c(1-0.1, 1-0.2)
) %>%
  mutate(post=prior*likelihood/sum(prior*likelihood))

df.pandas
```
Re-tweaking the priors and the likelihood gives us the right answer.

## 2h4 

A common boast of Bayesian statisticians is that Bayesian inference makes it 
easy to use all of  the data, even if the data are of different types. 

So suppose now that a veterinarian comes along who has a new genetic test that 
she claims can  identify the species of our mother panda. But the test, like all 
tests, is imperfect. This is the information you have about the test:

* The probability it correctly identifies a species A panda is 0.8.
* The probability it correctly identifies a species B panda is 0.65.

The vet administers the test to your panda and tells you that the test is 
positive for species A. 

First, ignore your previous information from the births and compute the 
posterior probability that your panda is species A. Then redo your calculation, 
now using the birth data as well.

```{r}
df.pandas = data.frame(
  panda_type=c("A", "B"),
  prior=c(1/2, 1/2),
  likelihood=c(0.8, 1-0.65)
) %>%
  mutate(post=prior*likelihood/sum(prior*likelihood))

df.pandas
```

```{r}
df.pandas = data.frame(
  panda_type=c("A", "B"),
  prior=c(1/2, 1/2),
  likelihood=c(0.8, 1-0.65)
) %>%
  mutate(post_test=prior*likelihood/sum(prior*likelihood),
         like_twins=c(0.1, 0.2),
         post_test_twins=post_test*like_twins/sum(post_test*like_twins),
         like_single=1-like_twins,
         post_test_twins_single=post_test_twins*like_single/
           sum(post_test_twins*like_single)
         )

df.pandas
```
